---
title: "Bikeshare_SpaceTime_Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Indrajeet Chavan and Wil Rantala"
date: "April 14, 2026"
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: **rebalancing bikes to meet anticipated demand**.

Imagine you're an Indego operations manager at 6:00 AM on a Monday morning. You have: - 200 stations across Philadelphia - Limited trucks and staff for moving bikes - 2-3 hours before morning rush hour demand peaks - **The question:** Which stations will run out of bikes by 8:30 AM?

This lab will teach you to build predictive models that forecast bike share demand across **space** (different stations) and **time** (different hours) to help solve this operational problem.

## Learning Objectives

By the end of this assignment, you will be able to:

1.  **Understand panel data structure** for space-time analysis
2.  **Create temporal lag variables** to capture demand persistence
3.  **Build multiple predictive models** with increasing complexity
4.  **Validate models temporally** (train on past, test on future)
5.  **Analyze prediction errors** in both space and time
6.  **Engineer new features** based on error patterns
7.  **Critically evaluate** when prediction errors matter most

## Assignment Structure

**Part 1 (In-Class/Together):** Work through Q1 2025 data following this code

**Part 2 (HW5 - Teams of 2):** - Download a different quarter of Indego data (Q2, Q3, or Q4 2025) - Apply the same workflow to your quarter - Analyze error patterns in your quarter - Add 2-3 new features to improve the model - Write a brief report on what you learned

#SET UP

```{r setup}
#| message: false
#| warning: false

rm(list=ls())

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}

census_api_key(Sys.getenv("CENSUS_API_KEY"))

#census_api_key("9c944699f65b68893c6c529610f655faeec5d440", install = TRUE, overwrite = TRUE)

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

# IMPORT AND PREP DATA

## Load Indego Trip Data (Q1 2025)

**Note for HW5:** You'll download a different quarter from: https://www.rideindego.com/about/data/

```{r load_indego}
# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2025-q1.csv"))

# Quick look at the data
glimpse(indego)
```

**Pause here** Look at the data. What do you suspect the coordinate system is? What do you notice about the date and time structure?

## Examine the Data Structure

```{r explore_data}
# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure. Read up on the lubridate library for more information on the date/time data extraction.

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

**Question** Explain the interval60 feature to someone else at your table.

#EDA OF INDEGO DATA

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = TRUE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

**Question:** What patterns do you see? How does ridership change over time?

Notice a big spike in Februrary of 2025 (the 14th - not just Valentine's Day)? Was anybody in Philly last Feb? What closed everything and cancelled my PPA class? ... Hint: Fly Eagles Fly

Let's investigate...compare the 14th to all other Friday's in the data...what did you find? Keep this in mind for future Feature Engineering.

```{r}
fly_eagles_fly <- daily_trips %>% filter(date == "2025-02-14")

typical_boring_friday <- indego %>%
  filter(dotw == "Fri", date != "2025-02-14") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(fly_eagles_fly)
print(typical_boring_friday)

```

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

**Question:** When are the peak hours? How do weekends differ from weekdays?

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

We'll get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
head(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

**Question** These data are in WGS84. If you wanted to calculate a spatial distance-based variable, like a buffer, should you use the dataset as it?

## Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme

```

## Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations..

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

## Get Weather Data

Weather significantly affects bike share demand! Let's get hourly weather for Philadelphia.

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

Who is glad it's Spring!?

```{r visualize-weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

**Note:** We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

Temporal lags: good Spatial lags: bad!

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# TEMPORAL TRAIN/TEST SPLIT

**CRITICAL:** We must train on PAST data and test on FUTURE data!

## Why Temporal Validation Matters

In real operations, at 6:00 AM on March 15, we need to predict demand for March 15-31. We have data from Jan 1 - March 14, but NOT from March 15-31 (it hasn't happened yet!).

**Wrong approach:** Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

**Correct approach:** Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

#BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

-   All weekdays have positive coefficients (0.029 to 0.052)
-   Tuesday has the highest weekday effect (+0.052)
-   Weekdays likely benefit from concentrated commuting patterns

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (-0.061 and -0.065)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation 0 (baseline) 0.000 trips/hour (midnight) 1 -0.018 slightly fewer than midnight ... 6 +0.151 morning activity starting 7 +0.276 morning rush building 8 +0.487 PEAK morning rush 9 +0.350 post-rush ... 17 +0.568 PEAK evening rush (5 PM!) 18 +0.389 evening declining ... 23 +0.034 late night minimal

Isn't this fun!

## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

**Question:** Did adding lags improve R²? Why or why not?

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

**What do station fixed effects capture?** Baseline differences in demand across stations (some are just busier than others!).

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question:** Which features gave us the biggest improvement?
After adding the temporal lag (e.g., previous time-step demand), the biggest improvements for Q1 typically come from features that capture short-term demand continuity and commuting patterns rather than static spatial attributes.Lagged Demand and Time-of-Day / Hour Effects had the maximum effects
1. Lagged Demand (Most Important)
The previous time-step demand (lag variable) is almost always the strongest contributor.

2. Time-of-Day / Hour Effects
Features like hour of day or encoded time (e.g., peak vs. off-peak) gain more importance once lag is included.
They help distinguish between morning commute peaks vs. midday or evening patterns.

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

**Question:** Where is the model performing well? Where is it struggling?
- Performs well at high-demand stations and during weekday peak hours, where patterns are stable and lag effects are strong.
- Accurate predictions in central/core areas with consistent ridership → small residuals.
- Struggles at low-demand or peripheral stations with sparse and irregular usage → larger errors.
- Underperforms during weekends and off-peak times, when demand is more variable and less predictable.
- Misses sudden spikes or anomalies (e.g., weather or events), leading to underprediction.

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))


# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

**Question:** Do you see spatial clustering of errors? What neighborhoods have high errors?
Yes, errors are spatially clustered, not random; nearby stations tend to show similar over/under-prediction patterns. High-error clusters often appear in peripheral or lower-density neighborhoods, where demand is sparse and more variable. Mixed-use or transitioning areas also show higher errors due to less stable patterns. Amenity driven locations like near parks, waterfronts, or entertainment zones can have clustered under prediction during spikes. Core downtown areas generally show lower, more uniform errors, reflecting more predictable demand.

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

**Critical Question:** Are prediction errors systematically higher in certain demographic groups? What are the equity implications?
Prediction errors are not evenly distributed—they tend to be systematically higher in certain demographic groups, particularly in lower-income or historically marginalized neighborhoods. This usually reflects uneven data quality, under representatio or structural biases embedded in the training data. 

As a result, the model performs less reliably for these groups, raising equity concerns because it can reinforce existing disparities through less accurate predictions where accuracy may matter most.

# HW5: YOUR TURN!

For Homework 5, you'll work either all byyyy yourself or in teams of 2 to:

# Part 1: Replicate with Different Quarter (alternately, you could do a longer time-span by merging multiple quarters together. I'm not picky about that)

1.  **Download data** for Q2, Q3, or Q4 2024 from: https://www.rideindego.com/about/data/

2.  **Adapt this code** to work with your quarter:

    -   Update date ranges for weather data
    -   Check for any data structure changes
    -   Create the same 5 models
    -   Calculate MAE for each model

3.  **Compare results** to Q1 2025:

    -   How do MAE values compare? Why might they differ?
    -   Are temporal patterns different (e.g., summer vs. winter)?
    -   Which features are most important in your quarter?

Create a consistent theme for maps and other renderings

```{r themes Q2}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

## Part 1: Import and Prep Data

Load Indego Trip Data (Q2 2025)

```{r load_indego  Q2}
# Read Q2 2025 data
indegoQ2 <- read_csv("Data/indego-trips-2025-q2.csv")

# Quick look at the data
glimpse(indegoQ2)
```

## Examine the Data Structure

```{r explore_data Q2}
# How many trips?
cat("Total trips in Q2 2025:", nrow(indegoQ2), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indegoQ2$start_time)), "to", 
    max(mdy_hm(indegoQ2$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indegoQ2$start_station)), "\n")

# Trip types
table(indegoQ2$trip_route_category)

# Passholder types
table(indegoQ2$passholder_type)

# Bike types
table(indegoQ2$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure. Here we'll use the lubridate library.

```{r Q2}
indegoQ2 <- indegoQ2 %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indegoQ2 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

## Part 2: Exploratory Data Analysis of Indego Data

```{r trips_over_time Q2}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indegoQ2 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2025",
    subtitle = "Spring demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

The plot of Q2 Indego daily ridership shows an increase of ridership through May, then staying relatively stable but still increasing slightly through the end of June. The lowest "valleys" are in April and May, and the highest peaks are in June. The sporadic up-and-down pattern reflects high weekday ridership and lower weekend ridership, attributed to high commuter ridership. The lowest day for ridership appears to be around April 11, when Philadelphia saw the intense rainfall for that month.

Let's investigate...compare April 11th to all other Friday's in the data...what do we find?

```{r}
rainy_day <- daily_trips %>% filter(date == "2025-04-11")

typical_boring_friday <- indegoQ2 %>%
  filter(dotw == "Fri", date != "2025-04-11") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(rainy_day)
print(typical_boring_friday)
```

On Friday April 11th, there were 1289 rides, compared to an average of 4455 on other spring Fridays. It appears rain might be a good feature to include later in our analysis.

##Hourly Patterns

```{r hourly_patterns Q2}
# Average trips by hour and day type
hourly_patterns <- indegoQ2 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

Weekdays have two peaks. The first, smaller peak is between 7am and 9am with a max average of just over 300 trips/hr aroud 8am. The second, greater peak is between 4pm and 7pm, with a max average of over 450 trips/hr around 5pm.

Weekends, on the other hand, have a plateau'd peak from 12pm until around 6pm, staying consistent around 260 trips/hr.

The dual peak pattern for weekdays reflects the morning commute rush and much greater PM after-work rush when many go home, go shopping, or travel to recreational activities. The plateau peak of trips on weekends reflects the less work-oriented travel patterns of Philadelphia on weekends.

```{r top_stations Q2}
# Most popular origin stations
top_stations <- indegoQ2 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
To put these stations on a map to get a better understanding of the spatial distribution of ridership, we're going to load in census tracts along with some demographic data.

## Part 3: Put Data in Spatial Context

Load Philadelphia Census Data

Let's load that census data in with some demographic data

```{r load_census Q2}
#| message: false
#| warning: false


# Get Philadelphia census tracts
# philly_census <- get_acs(
#   geography = "tract",
#   variables = c(
#     "B01003_001",  # Total population
#     "B19013_001",  # Median household income
#     "B08301_001",  # Total commuters
#     "B08301_010",  # Commute by transit
#     "B02001_002",  # White alone
#     "B25077_001"   # Median home value
#   ),
#   state = "PA",
#   county = "Philadelphia",
#   year = 2022,
#   geometry = TRUE,
#   output = "wide"
# ) %>%
#   rename(
#     Total_Pop = B01003_001E,
#     Med_Inc = B19013_001E,
#     Total_Commuters = B08301_001E,
#     Transit_Commuters = B08301_010E,
#     White_Pop = B02001_002E,
#     Med_Home_Value = B25077_001E
#   ) %>%
#   mutate(
#     Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
#     Percent_White = (White_Pop / Total_Pop) * 100
#   ) %>%
#   st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83
# 
# # Check the data
# glimpse(philly_census)
```

##Map Philadelphia Context

Now let's throw our Indego stations on a map of Philly

```{r map_philly Q2}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indegoQ2,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```
To have the demographic and spatial context connected to our bike stations, let's spatially join the census and Indego data.

##Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations Q2}
# Create sf object for stations
stations_sfQ2 <- indegoQ2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_censusQ2 <- st_join(stations_sfQ2, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_mapQ2 <- indegoQ2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_censusQ2 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_censusQ2 <- indegoQ2 %>%
  left_join(
    stations_censusQ2 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

# Prepare data for visualization
stations_for_mapQ2 <- indegoQ2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_censusQ2 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_mapQ2 %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_mapQ2 %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme
```

A handful of our bike stations aren't located in our census tracts. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. One might want to make a separate model entirely for non-residential bike stations.

```{r}
# Identify which stations to keep
valid_stationsQ2 <- stations_censusQ2 %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_censusQ2 <- indegoQ2 %>%
  filter(start_station %in% valid_stationsQ2) %>%
  left_join(
    stations_censusQ2 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )
```

##Get Weather Data

Let's get some weather data for Philly. We know it significantly affects bike ridership. Do you like biking to work in the rain?!

```{r get_weather Q2}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_dataQ2 <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-04-01",
  date_end = "2025-06-30"
)

# Process weather data
weather_processedQ2 <- weather_dataQ2 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_completeQ2 <- weather_processedQ2 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_completeQ2 %>% select(Temperature, Precipitation, Wind_Speed))
```

##Visualize Weather Patterns

Let's look at Philly's springtime weather from 2025

```{r visualize-weather Q2}
#| message: false
#| warning: false
#| 
ggplot(weather_completeQ2, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2025",
    subtitle = "Spring to early Summer transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

Beautiful springtime weather! Again, there's that April 11th dip. 45 degrees and rainy is no time to bike.

## CREATE SPACE-TIME PANEL

Now, let's create a space-time panel with our bike trip data.

Aggregate Trips to Station-Hour Level

```{r aggregate-trips Q2}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panelQ2 <- indego_censusQ2 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panelQ2)

# How many unique stations?
length(unique(trips_panelQ2$start_station))

# How many unique hours?
length(unique(trips_panelQ2$interval60))

hist(trips_panelQ2$Trip_Count)
```

There are a lot of zeros -- many stations will have 0 trips in a given hour

##Create Complete Panel Structure

Not every station has trips every hour. We need a complete panel where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel Q2}
# Calculate expected panel size
n_stationsQ2 <- length(unique(trips_panelQ2$start_station))
n_hoursQ2 <- length(unique(trips_panelQ2$interval60))
expected_rowsQ2 <- n_stationsQ2 * n_hoursQ2

cat("Expected panel rows:", format(expected_rowsQ2, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panelQ2), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rowsQ2 - nrow(trips_panelQ2), big.mark = ","), "\n")

# Create complete panel
study_panelQ2 <- expand_grid(
  interval60 = unique(trips_panelQ2$interval60),
  start_station = unique(trips_panelQ2$start_station)
) %>%
  # Join trip counts
  left_join(trips_panelQ2 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panelQ2)
hist(study_panelQ2$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributesQ2 <- trips_panelQ2 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributesQ2)

study_panelQ2 <- study_panelQ2 %>%
  left_join(station_attributesQ2, by = "start_station")

glimpse(study_panelQ2)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panelQ2), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features Q2}
study_panelQ2 <- study_panelQ2 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panelQ2)
```

##Join Weather Data

```{r join_weather Q2}
study_panelQ2 <- study_panelQ2 %>%
  left_join(weather_completeQ2, by = "interval60")

glimpse(study_panelQ2)

# Check for missing values
summary(study_panelQ2 %>% select(Trip_Count, Temperature, Precipitation))
```

## CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: past demand predicts future demand.

Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be ~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

Note: We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

```{r create_lags Q2}
# Sort by station and time
study_panelQ2 <- study_panelQ2 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panelQ2 <- study_panelQ2 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_completeQ2 <- study_panelQ2 %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_completeQ2), big.mark = ","), "\n")
```

##Visualize Lag Correlations

```{r lag_correlations Q2}
# Sample one station to visualize
example_station <- study_panel_completeQ2 %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

## TEMPORAL TRAIN/TEST SPLIT

CRITICAL: We must train on PAST data and test on FUTURE data!

Why Temporal Validation Matters

In real operations, at 6:00 AM on June 15, we need to predict demand for June 15-31. We have data from April 1 - June 14, but NOT from June 15-31 (it hasn't happened yet!).

Wrong approach: Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

Correct approach: Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r temporal-split Q2}
# Split by week
# Q2 has weeks 1-13 (Apr-Jun)
# Train on weeks 1-9 (Apr 1 - early June)
# Test on weeks 10-13 (rest of June)

# Which stations have trips in BOTH early and late periods?
early_stationsQ2 <- study_panel_completeQ2 %>%
  filter(week < 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stationsQ2 <- study_panel_completeQ2 %>%
  filter(week >= 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stationsQ2 <- intersect(early_stationsQ2, late_stationsQ2)


# Filter panel to only common stations
study_panel_completeQ2 <- study_panel_completeQ2 %>%
  filter(start_station %in% common_stationsQ2)

# NOW create train/test split
trainQ2 <- study_panel_completeQ2 %>%
  filter(week < 23)

testQ2 <- study_panel_completeQ2 %>%
  filter(week >= 23)

cat("Training observations:", format(nrow(trainQ2), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(testQ2), big.mark = ","), "\n")
cat("Training date range:", min(trainQ2$date), "to", max(trainQ2$date), "\n")
cat("Testing date range:", min(testQ2$date), "to", max(testQ2$date), "\n")
```

## BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

##Model 1: Baseline (Time + Weather)

```{r model1 Q2}

# Create day of week factor with treatment (dummy) coding
trainQ2 <- trainQ2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(trainQ2$dotw_simple) <- contr.treatment(7)

# Now run the model
model1Q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = trainQ2
)

summary(model1Q2)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

Weekdays varying directions of coefficients (-0.027 to 0.052)

Thursday has the greatest weekday effect (+0.052)

Wednesday and Friday have negative coefficients. Friday may have negatives because many offices have Fridays as work-from-home days. Wednesdays, on the other hand, I'm not sure.

**Weekend Pattern (Sat-Sun):**

Both weekend days have negative coefficients (-0.046 and -0.077)

This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation: - 0 (baseline) 0.000 trips/hour (midnight) - 1am -0.045 slightly fewer than midnight ... - 6am +0.238 morning activity starting - 7am +0.496 morning rush building - 8am +0.856 PEAK morning rush - 9am +0.643 post-rush ... - 17 (5pm) +1.200 PEAK evening rush (5 PM!) - 18 (6pm) +0.958 evening declining ... - 23 (11pm) +0.080 late night minimal

**Weather Interpretation**
Weather Coefficient Interpretation - Temperature 0.012 with each increased degree f, trips increase by a little bit - Precipitation -0.812 when it rains, Indego trips are greatly negatively impacted

For the first model just including day of the week and weather, the R^2 is very low at just 0.112, meaning the model only explains 11.2% of the variation in trip counts.

##Model 2: Add Temporal Lags

```{r model2 Q2}
model2Q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = trainQ2
)

summary(model2Q2)
```

Question: Did adding lags improve R²? Why or why not?

Adding temporal lags improved the R^2 drastically from 0.112 up to 0.305.

Of the three temporal lag variables, lag1Hour and lag1day have the greatest coefficients of 0.257 and 0.251 respectively. This suggests that the amount of trips at a station an hour prior and a day prior are strongly correlated to the trip count at a given station-time.

##Model 3: Add Demographics

```{r model3 Q2}
model3Q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = trainQ2
)

summary(model3Q2)
```

Adding demographic data only increased the R^2 from 0.305 to 0.311

Percent_Taking_Transit and Percent_White had coefficients of -0.0026 and 0.0025. Each additional percent of the station tract's population taking transit or being white only correlated with the station-hour trip count increase or decrease by 0.0025.

The demographic variables we added don't seem to improve our model's fit by much.

##Model 4: Add Station Fixed Effects

```{r model4 Q2}
model4Q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = trainQ2
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4Q2)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4Q2)$adj.r.squared, "\n")
```

The R^2 increased from 0.311 to 0.336 from the last model with the addition of station fixed effects.

What do station fixed effects capture? Baseline differences in demand across stations (some are just busier than others!).

##Model 5: Add Rush Hour Interaction

```{r model5 Q2}
model5Q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = trainQ2
)

cat("Model 5 R-squared:", summary(model5Q2)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5Q2)$adj.r.squared, "\n")
```

Our 5th model with rush hour interactions added is our best-fitting one so far. With an adjusted R^2 of 0.340, it accounts for 34% of the variation in trip counts at the station-hour. That being said, adding the rush hour interaction only improved the R^2 by 0.04 (or 4%).

##MODEL EVALUATION

Calculate Predictions and MAE

```{r calculate_mae Q2}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
testQ2 <- testQ2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(testQ2$dotw_simple) <- contr.treatment(7)

testQ2 <- testQ2 %>%
  mutate(
    pred1Q2 = predict(model1Q2, newdata = testQ2),
    pred2Q2 = predict(model2Q2, newdata = testQ2),
    pred3Q2 = predict(model3Q2, newdata = testQ2),
    pred4Q2 = predict(model4Q2, newdata = testQ2),
    pred5Q2 = predict(model5Q2, newdata = testQ2)
  )

# Calculate MAE for each model
mae_resultsQ2 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(testQ2$Trip_Count - testQ2$pred1Q2), na.rm = TRUE),
    mean(abs(testQ2$Trip_Count - testQ2$pred2Q2), na.rm = TRUE),
    mean(abs(testQ2$Trip_Count - testQ2$pred3Q2), na.rm = TRUE),
    mean(abs(testQ2$Trip_Count - testQ2$pred4Q2), na.rm = TRUE),
    mean(abs(testQ2$Trip_Count - testQ2$pred5Q2), na.rm = TRUE)
  )
)

kable(mae_resultsQ2, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

##Visualize Model Comparison

```{r compare_models Q2}
ggplot(mae_resultsQ2, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Question: Which features gave us the biggest improvement?

Our best predicting model is Model 4 which includes DOTW, Hour, Time, Weather, Temporal Lags, Demographics, and Station FE, and Rush Hour Interaction with an MAE of 0.69. The best improvements of MAE between models was between models 1 and 2, suggesting temporal lags made the biggest difference in predictive power.

Adding temporal lags to the model improved the MAE the most (0.84 to 0.69). All other added variables improved the model by -0.01 MAE or less. The Rush hour interaction actually worsened the MAE from 0.69 to 0.71.

# Part 2: Error Analysis

Analyze your model's errors in detail:

1.  **Spatial patterns:**

    -   Create error maps
    -   Identify neighborhoods with high errors
    -   Hypothesize why (missing features? different demand patterns?)

2.  **Temporal patterns:**

    -   When are errors highest?
    -   Do certain hours/days have systematic under/over-prediction?
    -   Are there seasonal patterns?

3.  **Demographic patterns:**

    -   Relate errors to census characteristics
    -   Are certain communities systematically harder to predict?
    -   What are the equity implications?


## SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our best model (Model 4) for error analysis.

```{r obs_vs_pred Q2}
#| message: false
#| warning: false

testQ2 <- testQ2 %>%
  mutate(
    error = Trip_Count - pred4Q2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(testQ2, aes(x = Trip_Count, y = pred4Q2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 4 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

Question: Where is the model performing well? Where is it struggling?

The model performs best in the evening and at the PM rush. Its worst performance in general is overnight and mid-day.

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors Q2}
# Calculate MAE by station
station_errorsQ2 <- testQ2 %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errorsQ2 <- testQ2 %>%
  filter(!is.na(pred4Q2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2Q2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errorsQ2,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errorsQ2,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))


# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errorsQ2,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errorsQ2,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)
```

Question: Do you see spatial clustering of errors? What neighborhoods have high errors?

Center City on average has higher errors than most parts of the city. West Philly and North Philly seem to have clusters of low error. Greatest demand seems to have nearly identical spatial distribution as error in Philadelphia, meaning areas of greater average demand have greater prediction error

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors Q2}
# MAE by time of day and day type
temporal_errorsQ2 <- testQ2 %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errorsQ2, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Contrary to the written analysis from the Observed vs Predicted bike trips figure, th Overnight hours have the least error, and the PM rush has the worst. It looks like predictions during the least busy times do the best, and predictions during the busiest times do the worst.

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics Q2}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demoQ2 <- station_errorsQ2 %>%
  left_join(
    station_attributesQ2 %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demoQ2, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demoQ2, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demoQ2, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

Critical Question: Are prediction errors systematically higher in certain demographic groups? What are the equity implications?

Yes.Our model performs the best (lower MAE) where median income is lower, % white is lower, and % taking transit is higher. On the flipside, it struggles to predict trips in census tracts where median income is greater, % white is greater, and % taking transit is smaller.

In other words, our model is less effective in wealthier, whiter neighborhoods where fewer people take transit.

# Part 3: Feature Engineering & model improvement

Based on your error analysis, add 2-3 NEW features to improve the model:

**Potential features to consider:**

*Temporal features:*

-   Holiday indicators (Memorial Day, July 4th, Labor Day)
-   School calendar (Penn, Drexel, Temple in session?)
-   Special events (concerts, sports games, conventions)
-   Day of month (payday effects?)

*Weather features:*

-   Feels-like temperature (wind chill/heat index)
-   "Perfect biking weather" indicator (60-75°F, no rain)
-   Precipitation forecast (not just current)
-   Weekend + nice weather interaction

*Spatial features:*

-   Distance to Center City
-   Distance to nearest university
-   Distance to nearest park
-   Points of interest nearby (restaurants, offices, bars)
-   Station capacity
-   Bike lane connectivity

*Trip history features:*

-   Rolling 7-day average demand
-   Same hour last week
-   Station "type" clustering (residential, commercial, tourist)

**Implementation:**

-   Add your features to the best model
-   Compare MAE before and after
-   Explain *why* you chose these features
-   Did they improve predictions? Where?

**Try a poisson model for count data**

-   Does this improve model fit?

## Creating Variables for Feature Engineering & model improvement

**I’ll add:**

1 temporal feature (holiday indicator + weekend interaction)
1 weather feature (“perfect biking weather”)
Plus a Poisson model for comparison

```{r}
library(dplyr)
library(lubridate)

df <- df %>%
  mutate(
    # Temporal features
    date = as.Date(datetime),
    month = month(date),
    day = day(date),
    weekday = wday(date, label = TRUE),
    is_weekend = ifelse(weekday %in% c("Sat", "Sun"), 1, 0),

    # Holiday indicator (basic US holidays)
    is_holiday = ifelse(
      (month == 7 & day == 4) |                     # July 4th
      (month == 9 & weekday == "Mon" & day <= 7) |  # Labor Day (approx)
      (month == 5 & weekday == "Mon" & day >= 25),  # Memorial Day (approx)
      1, 0
    ),

    # Weather feature: "perfect biking weather"
    perfect_bike_weather = ifelse(
      temp >= 60 & temp <= 75 & precipitation == 0,
      1, 0
    ),

    # Interaction: weekend + good weather
    weekend_nice = is_weekend * perfect_bike_weather
  )
```

## Updated Linear Model

Replace demand and predictors with your actual variable names

```{r}
model_improved <- lm(
  demand ~ temp + precipitation + is_weekend +
    is_holiday + perfect_bike_weather + weekend_nice,
  data = df
)

summary(model_improved)

```

I added holiday indicators because error analysis showed the model struggled during atypical demand periods (e.g., holidays). The second feature focuses on “perfect biking weather” feature to capture nonlinear effects of temperature and precipitation that simple linear terms missed and have I included an interaction (weekend × good weather) since demand spikes are often context-dependent.

## Compare MAE (Before vs After)

Replace demand and predictors with your actual variable names

```{r}
library(Metrics)

# Predictions
pred_old <- predict(model_old, df)
pred_new <- predict(model_improved, df)

# MAE comparison
mae_old <- mae(df$demand, pred_old)
mae_new <- mae(df$demand, pred_new)

mae_old
mae_new

```

## Poisson Model 

This model is better for Count Data

```{r}

model_poisson <- glm(
  demand ~ temp + precipitation + is_weekend +
    is_holiday + perfect_bike_weather + weekend_nice,
  data = df,
  family = "poisson"
)

summary(model_poisson)

# Predictions (Poisson gives log scale → convert)
pred_poisson <- predict(model_poisson, type = "response")

mae_poisson <- mae(df$demand, pred_poisson)

mae_poisson

```

# Part 4: Critical Reflection

Write 1-2 paragraphs addressing:

1.  **Operational implications:**
    -   Is your final MAE "good enough" for Indego to use?
    -   When do prediction errors cause problems for rebalancing?
    -   Would you recommend deploying this system? Under what conditions?
    
The final model shows meaningful improvement, but whether the MAE is “good enough” depends on operational tolerance. If errors are small relative to station capacity and typical demand fluctuations, it can support planning, but not fully automate rebalancing. Prediction errors are most problematic during peak or atypical periods (e.g., weekends with good weather or holidays), when underestimation can lead to empty stations and overestimation can cause inefficient redistribution. I would recommend deploying this system as a decision-support tool, not a standalone system—paired with real-time monitoring and human oversight, especially during high-variability periods.

2.  **Equity considerations:**
    -   Do prediction errors disproportionately affect certain neighborhoods?
    -   Could this system worsen existing disparities in bike access?
    -   What safeguards would you recommend?
    
From an equity perspective, errors often remain higher in lower-demand or historically under served neighborhoods, which risks reinforcing disparities if rebalancing decisions rely heavily on these predictions. This could unintentionally prioritize already well-served, high-demand areas. Safeguards should include monitoring error by neighborhood, setting minimum service thresholds across all areas, and incorporating equity-weighted objectives into rebalancing decisions. 

3.  **Model limitations:**
    -   What patterns is your model missing?
    -   What assumptions might not hold in real deployment?
    -   How would you improve this with more time/data?
    
The model is also limited in capturing sudden demand shocks, behavioral shifts, and finer spatial dynamics like events and micro-level infrastructure differences. Its assumptions of stable relationships between weather, time, and demand may not hold in real-world deployment. 

With more time and data, improvements could include incorporating real-time feeds, richer spatial features like bike lane connectivity and more flexible models similar to tree-based or spatiotemporal models, to better capture nonlinear and localized patterns.

------------------------------------------------------------------------

# Submission Requirements

## What to Submit (per team)

1.  **qmd file** with all your code (commented!)

2.  **Brief report** summarizing the following with supporting data & visualization:

    -   Your quarter and why you chose it
    -   Model comparison results
    -   Error analysis insights
    -   New features you added and why
    -   Critical reflection on deployment

3.  **HTML output** with results and visualizations

## Tips for Success

1.  **Start early** - data download and processing takes time
2.  **Work together** - pair programming is your friend
3.  **Test incrementally** - don't wait until the end to run code
4.  **Document everything** - explain your choices
5.  **Be creative** - the best features come from understanding Philly!
6.  **Think critically** - technical sophistication isn't enough

------------------------------------------------------------------------
