---
title: "Midterm Challenge: Philadelphia Housing Price Prediction (Technical Appendix)"
subtitle: "MUSA 5080 - Public Policy Analytics"
author: "Indrajeet Chavan, Wil Rantala, Alajandro Duque, Umair Saad"
date: "March 16, 2026"
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

#Overview

**The Challenge:**
Build a predictive model that city officials would actually want to use, using only the techniques we’ve covered this semester.

**The Stakes:**
Your presentation will be evaluated by a panel of judges (your professor + a surprise guest) who will grill you on your methods, assumptions, and recommendations. You need to demonstrate not just technical competence, but deep understanding of what you’ve built and why it matters.

## Phase 1: Data Cleaning

#### Step 1: Load Libraries

```{r}
#Clearing previous data
rm(list=ls())

library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(lubridate)
library(janitor)
library(knitr)
library(ggplot2)
library(dplyr)
library(janitor)

options(tigris_use_cache = TRUE)

```

#### Step 2: Load & Inspect Bus Ridership

```{r}

#Load ridership data
bus_rider_2022 <- st_read("Data/Fall_2022_Stop_Summary_(Bus).csv") %>%
  clean_names() %>%
  select(-source,-gis_db_id) 

bus_rider_2023 <- st_read("Data/Fall_2023_Stop_Summary_(Bus).csv") %>%
  clean_names() %>%
  select(-gisdbid)

bus_rider_2024 <- st_read("Data/Fall_2024_Stop_Summary_(Bus).csv") %>%
  clean_names()

bus_rider_2025 <- st_read("Data/Fall_2025_Stop_Summary_(Bus).csv") %>%
  clean_names()

# Standardize column names
colnames(bus_rider_2022) <- colnames(bus_rider_2023) <- 
colnames(bus_rider_2024) <- colnames(bus_rider_2025) <- 
c("x","y","fid","mode","route","direction","route_1","stop",
  "lat","lon","season","week_on","week_off",
  "sat_on","sat_off","sun_on","sun_off")

# Combine
bus_rider <- bind_rows(
  bus_rider_2022,
  bus_rider_2023,
  bus_rider_2024,
  bus_rider_2025
)

```

#### Step 3: Clean + Create Target Variable

```{r}
# Load Philadelphia boundary
bus_rider <- bus_rider %>%
  mutate(
    week_on = as.numeric(week_on),
    sat_on  = as.numeric(sat_on),
    sun_on  = as.numeric(sun_on),
    total_riders = week_on + sat_on + sun_on
  )

```

#### Step 4: Convert to Spatial

```{r}
# Load Philadelphia boundary
bus_rider <- bus_rider %>%
  mutate(
    x = as.numeric(x),
    y = as.numeric(y),
    total_riders = week_on + sat_on + sun_on
  ) %>%
  filter(!is.na(x), !is.na(y), total_riders > 0)

bus_rider <- bus_rider %>%
  filter(!is.na(total_riders))

bus_rider_sf <- bus_rider %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_transform(2272)

```

#### Step 5: Clip to Philadelphia

```{r}
# Load Philadelphia boundary
philadelphia <- get_acs(
  geography = "county",
  variables = "B01003_001",
  state = "PA",
  county = "Philadelphia",
  geometry = TRUE,
  year = 2023
) %>%
  st_transform(2272)

bus_rider_sf <- st_intersection(bus_rider_sf, philadelphia)
```

#### Step 6: Distance to Nearest Transit Stop

```{r}

neighbors <- st_is_within_distance(bus_rider_sf, dist = 500) # 500 meters

bus_rider_sf$neighbor_count <- lengths(neighbors)

```

#### Step 7: Load the roads

```{r}
roads <- st_read("Data/PaLocalRoads2024_03/PaLocalRoads2024_03.shp") %>%
  st_transform(2272)

```