---
title: "Transit Prediction Model- Technical Appendix"
subtitle: "MUSA 5080 - Public Policy Analytics"
author: "Indrajeet Chavan, Wil Rantala, Alajandro Duque, Umair Saad"
date: "March 16, 2026"
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

#Overview

**The Challenge:**
Build a predictive model that city officials would actually want to use, using only the techniques we’ve covered this semester.

**The Stakes:**
Your presentation will be evaluated by a panel of judges (your professor + a surprise guest) who will grill you on your methods, assumptions, and recommendations. You need to demonstrate not just technical competence, but deep understanding of what you’ve built and why it matters.

## Phase 1: Data Cleaning

#### Step 1: Load Libraries

```{r}
#Clearing previous data
rm(list=ls())

install.packages("lehdr")

library(lehdr)
library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(lubridate)
library(janitor)
library(knitr)
library(ggplot2)
library(dplyr)
library(janitor)
library(dplyr)
library(sf)

options(tigris_use_cache = TRUE)
```

#### Step 2: Load & Inspect Bus Ridership

```{r}

#Load ridership data
bus_rider_2022 <- st_read("Data/Fall_2022_Stop_Summary_(Bus).csv") %>%
  clean_names() %>%
  select(-source,-gis_db_id) 

bus_rider_2023 <- st_read("Data/Fall_2023_Stop_Summary_(Bus).csv") %>%
  clean_names() %>%
  select(-gisdbid)

bus_rider_2024 <- st_read("Data/Fall_2024_Stop_Summary_(Bus).csv") %>%
  clean_names()

bus_rider_2025 <- st_read("Data/Fall_2025_Stop_Summary_(Bus).csv") %>%
  clean_names()

# Standardize column names
colnames(bus_rider_2022) <- colnames(bus_rider_2023) <- 
colnames(bus_rider_2024) <- colnames(bus_rider_2025) <- 
c("x","y","fid","mode","route","direction","route_1","stop",
  "lat","lon","season","week_on","week_off",
  "sat_on","sat_off","sun_on","sun_off")

# Combine
bus_rider <- bind_rows(
  bus_rider_2022,
  bus_rider_2023,
  bus_rider_2024,
  bus_rider_2025
)

```

#### Step 3: Clean + Create Target Variable

```{r}

bus_rider <- bus_rider %>%
  mutate(
    week_on = as.numeric(week_on),
    sat_on  = as.numeric(sat_on),
    sun_on  = as.numeric(sun_on),
    total_riders = week_on + sat_on + sun_on
  ) %>%
  filter(!is.na(x), !is.na(y), total_riders > 0)

# Convert to sf
bus_rider_sf <- bus_rider %>%
  mutate(
    lat = as.numeric(lat),
    lon  = as.numeric(lon)
    ) %>%
  st_as_sf(coords = c("lon","lat"), crs = 4326) %>%
  st_transform(2272)

ggplot(bus_rider_sf) +
  geom_sf(color = "blue", size = 1) +
  theme_minimal() +
  labs(title = "Bus Stops")

```

#### Step 4: Create 1000 × 1000 ft Grid (your fishnet)

```{r}

# Philly boundary
philadelphia <- get_acs(
  geography = "county",
  variables = "B01003_001",
  state = "PA",
  county = "Philadelphia",
  geometry = TRUE,
  year = 2023
) %>%
  st_transform(2272)

# 1000 ft grid
philly_grid <- st_make_grid(philadelphia, cellsize = 1000, square = TRUE) %>%
  st_sf() %>%
  mutate(cell_id = row_number())

# Clip to city
philly_grid <- philly_grid[philadelphia, ]

# 3. Pull census tracts with raw counts
acs_vars <- c(
  total_pop     = "B01003_001",
  total_hh      = "B08201_001",
  no_vehicle_hh = "B08201_002",
  median_income = "B19013_001",
  poverty_denom = "B17001_001",
  poverty_count = "B17001_002",
  white         = "B03002_003",
  black         = "B03002_004",
  hispanic      = "B03002_012",
  asian         = "B03002_006"
)

tracts <- get_acs(
  geography = "tract",
  variables = acs_vars,
  state = "PA",
  county = "Philadelphia",
  geometry = TRUE,
  year = 2024,
  output = "wide"
) %>%
  st_transform(2272)

# 4. Interpolate counts onto grid
# 4. Interpolate — keep geometry so we can join spatially
grid_counts <- st_interpolate_aw(
  tracts %>% select(total_popE, poverty_countE, poverty_denomE,
                    no_vehicle_hhE, total_hhE,
                    whiteE, blackE, hispanicE, asianE, geometry),
  philly_grid,
  extensive = TRUE
)

grid_income <- st_interpolate_aw(
  tracts %>% select(median_incomeE, geometry),
  philly_grid,
  extensive = FALSE
) %>%
  rename(median_income = median_incomeE)

# 5. Spatial join back to grid (avoids row_number() misalignment)
philly_grid_census <- philly_grid %>%
  st_join(grid_counts %>% select(-geometry), join = st_equals) %>%
  st_join(grid_income  %>% select(median_income, geometry), join = st_equals) %>%
  mutate(
    cell_area_km2  = (1000 * 0.3048)^2 / 1e6,  # This is to convert it to km^2 from feet
    pop_density    = total_popE / cell_area_km2,
    pct_no_vehicle = no_vehicle_hhE / total_hhE,
    pct_poverty    = poverty_countE / poverty_denomE,
    pct_white      = whiteE / total_popE,
    pct_black      = blackE / total_popE,
    pct_hispanic   = hispanicE / total_popE,
    pct_asian      = asianE / total_popE,
  )
# Filter out any cells that have 0 population
philly_grid_census <- philly_grid_census %>%
  filter(!is.na(total_popE), total_popE > 0)

# We get some NAs for median_income, let's get neighboring cells' incomes and average them for those. On one pass of this function, there were still some NAs because the cells were using adjacent cells' NAs to try to fill in median income. For that reason, we're doing a for loop to repeat the function until there aren't any NAs.

fill_na_from_neighbors_iter <- function(grid, col, max_iter = 10) {
  vals <- grid[[col]]
  nb <- st_touches(grid)
  
  for (i in seq_len(max_iter)) {
    if (!any(is.na(vals))) break  # stop early if no NAs remain
    
    vals <- sapply(seq_along(vals), function(i) {
      if (!is.na(vals[i])) return(vals[i])
      neighbor_vals <- vals[nb[[i]]]
      neighbor_vals <- neighbor_vals[!is.na(neighbor_vals)]
      if (length(neighbor_vals) == 0) return(NA)
      mean(neighbor_vals)
    })
  }
  vals
}

philly_grid_census <- philly_grid_census %>%
  mutate(median_income = fill_na_from_neighbors_iter(., "median_income"))

# This still spits out a few NAs. Most likely, it's beacause some of the NA cells had NAs neighboring them.
```

#### Step 5: Visualizing the census grid

```{r}
#Visualizing the census grid
ggplot(philly_grid_census) +
  aes(fill = pop_density) +
  geom_sf(color = NA) +
  scale_fill_viridis_c(option = "magma", name = "Pop/km²") +
  labs(title = "Population Density by Grid Cell") +
  theme_void()
```

#### Step 6: Assign Bus Stops to Grid Cells

```{r}

# Spatial join: assign each bus stop to a grid cell
bus_rider_sf <- bus_rider_sf %>%
  select(-x, -y) 

bus_with_grid <- st_join(bus_rider_sf, philly_grid, join = st_intersects)

# Clip to city
bus_with_grid <- bus_with_grid[philadelphia, ]
  
# head(philly_grid)
# 
# sum(lengths(st_within(bus_rider_sf, philly_grid)) > 0)
# 
# plot.new()
# plot(st_geometry(philly_grid), border = "grey")
# plot(st_geometry(bus_rider_sf), add = TRUE, col = "red", pch = 16)
# 
# sum(st_is_empty(bus_rider_sf))
# 
# library(ggplot2)
# 

ggplot() +
  geom_sf(data = philly_grid, fill = NA, color = "grey") +
  geom_sf(data = bus_rider_sf, color = "red", size = 1) +
  theme_minimal()

ggplot() +
  geom_sf(data = bus_with_grid, color = "red", size = 1) +
  theme_minimal()

# sum(is.na(st_coordinates(bus_rider_sf)))
```

#### Step 7: Aggregate to Grid Level

```{r}

bus_with_grid <- bus_with_grid %>%
  mutate(
    coords = st_coordinates(geometry),
    stop_id = paste0(round(coords[,1], 2), "_", round(coords[,2], 2))
  )

bus_grid_summary <- bus_with_grid %>%
  st_drop_geometry() %>%
  group_by(cell_id) %>%
  summarise(
    total_riders = sum(total_riders, na.rm = TRUE),

    # unique stops (based on coordinates)
    unique_stops = n_distinct(stop_id),

    # unique routes in that grid cell
    unique_routes = n_distinct(route)
  )

```

#### Step 6: Join Back to Grid

```{r}

philly_grid_final <- philly_grid_census %>%
  left_join(bus_grid_summary, by = "cell_id") %>%
  mutate(
    total_riders = replace_na(total_riders, 0),
    unique_stops   = replace_na(unique_stops, 0),
    unique_routes   = replace_na(unique_routes, 0)
  )

```

#### Step 7: Loading Spatial Data: Roads

```{r}
#Loading the shape file 
roads <- st_read("Data/PaLocalRoads2024_03/PaLocalRoads2024_03.shp")

#Setting the CRS
roads <- st_transform(roads, st_crs(philly_grid))

```

#### Step 8: Create a binary road indicator

```{r}
#Loading the shape file 
road_indicator <- philly_grid %>%
  mutate(has_road = as.integer(lengths(st_intersects(., roads)) > 0)) %>%
  st_drop_geometry() %>%
  select(cell_id, has_road)

table(road_indicator$has_road, useNA = "ifany")

```

#### Step 9: Join categorical variable for roads to your final dataset

```{r}

philly_grid_final <- philly_grid_final %>%
  left_join(road_indicator, by = "cell_id") %>%
  mutate(has_road = replace_na(has_road, 0))

#Checking if it worked
table(philly_grid_final$has_road, useNA = "ifany")

#Making a dataset for road length in each grid cell
road_length <- st_intersection(philly_grid, roads) %>%
  mutate(length = st_length(geometry)) %>%
  st_drop_geometry() %>%
  group_by(cell_id) %>%
  summarise(road_length = sum(length))

```

#### Step 10: Mapping which grid cells have roads

```{r}
#Creating a map for mapping cells with roads
# ggplot(philly_grid_final) +
#   geom_sf(aes(fill = factor(has_road)), color = "grey70", size = 0.1) +
#   scale_fill_manual(
#     values = c("0" = "maroon", "1" = "darkblue"),
#     labels = c("0" = "No Road", "1" = "Road Present"),
#     name = "Road Presence"
#   ) +
#   theme_minimal() +
#   labs(title = "Grid Cells with Road Presence")

ggplot(philly_grid_final) +
  geom_sf(aes(fill = factor(has_road)), color = NA) +
  scale_fill_manual(values = c("0" = "darkgrey", "1" = "darkblue")) +
  theme_void() +
  labs(title = "Road Presence by Grid Cell")

```

#### Step 11: Load Data for Jobs

```{r}
# #Loading the shape file 
# jobs <- st_read("Data/otm_44d5b32c1de048ccb04fcbfc95b0a180/polygon_2023.shp")
# 
# #Setting the CRS
# jobs <- st_transform(jobs, st_crs(philly_grid))

# Download PA Workplace Area Characteristics (jobs by location)
pa_wac <- grab_lodes(
  state = "pa",
  year = 2021,
  lodes_type = "wac",
  job_type = "JT00",
  segment = "S000",
  state_part = "main",
  agg_geo = "block"   # ← IMPORTANT
)

philly_jobs <- pa_wac %>%
  filter(substr(w_geocode, 1, 5) == "42101")


blocks <- get_decennial(
  geography = "block",
  variables = "P1_001N",
  state = "PA",
  county = "Philadelphia",
  year = 2020,
  geometry = TRUE
) %>%
  st_transform(2272)

```

#### Step 12: Join Jobs to Blocks

```{r}
#Join jobs to blocks
philly_jobs_sf <- blocks %>%
  left_join(
    philly_jobs,
    by = c("GEOID" = "w_geocode")
  )

#Cleaning the data to keep what we need
philly_jobs_sf <- philly_jobs_sf %>%
  select(GEOID, C000, geometry) %>%
  rename(total_jobs = C000) %>%
  mutate(total_jobs = replace_na(total_jobs, 0))
```

#### Step 13: Aggregate Data to the grid

```{r}
# spatial join blocks to the grid
jobs_with_grid <- st_join(philly_jobs_sf, philly_grid, join = st_within)

# aggregate
jobs_grid <- jobs_with_grid %>%
  st_drop_geometry() %>%
  group_by(cell_id) %>%
  summarise(
    total_jobs = sum(total_jobs, na.rm = TRUE)
  )

#Adding to the final grid
philly_grid_final <- philly_grid_final %>%
  left_join(jobs_grid, by = "cell_id") %>%
  mutate(total_jobs = replace_na(total_jobs, 0))
```

#### Step 13: Load Data for Jobs

```{r}

philly_grid_final <- philly_grid_final %>%
  mutate(
    jobs_density = total_jobs / ((1000 * 0.3048)^2),
    jobs_to_pop_ratio = total_jobs / (total_popE + 1)
  )

```

#### Step 14: Mapping the Job density

```{r}
#Creating a visualization for total jobs

ggplot(philly_grid_final) +
  geom_sf(aes(fill = total_jobs), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    trans = "sqrt",
    limits = c(0, 1500),
    oob = scales::squish
  ) +
  theme_void() +
  labs(
    title = "Job Density by Grid Cell",
    fill = "Total Jobs"
  )
```
