---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: Indrajeet Chavan
date: January 30, 2026
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **Philadelphia/Pennsylvania Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup


```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("9c944699f65b68893c6c529610f655faeec5d440", install = TRUE, overwrite = TRUE)

# Choose your state for analysis - assign it to a variable called my_state
my_state <- "PA"
```

**State Selection:** I have chosen **Pennsylvania** for this analysis because: Philadelphia is my home. I am deeply interested in understanding the factors that contribute to the design of algorithmic systems used to identify communities in need of priority social service funding and outreach. This exercise also provides an opportunity to critically evaluate the quality of census data across the state and assess how data reliability may influence funding decisions that affect vulnerable populations. Additionally, I am excited to possibly contribute in unlocking some important funds for my people.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
PA_data <- get_acs(geography = "County", state = "PA", 
                   variables = c(medinc = "B19013_001", 
                                 totpop = "B01003_001"), #Median Income and Total Population
                   year = 2022, survey = "acs5", 
                   output = "wide")

PA_data

# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
PA <- PA_data %>%
  mutate(
    # Remove state name from county names
    county_name = str_remove(NAME, ", Pennsylvania"),
    # Remove "County" word
    county_name = str_remove(county_name, " County")
  )

# Display the first few rows
head(PA)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
PA_Reliability <- PA %>%
  mutate( moe_pct = (medincM/medincE)*100,
          reliability = case_when(
            moe_pct < 5 ~ "High Confidence",
            moe_pct >= 5 & moe_pct <= 10 ~ "Moderate Confidence",
            moe_pct > 10 ~ "Low Confidence"))

head(PA_Reliability)

# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
Reliability_Sum <- PA_Reliability %>%
  count(reliability) %>%
  mutate(percentage = round((n/sum(n)*100)))

Reliability_Sum
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
PA_Uncertainity <- PA_Reliability %>%
  arrange(desc(moe_pct)) %>%
  slice(1:5) %>%
  select(county_name, medincE, medincM, moe_pct, reliability)

# Format as table with kable() - include appropriate column names and caption
kable(PA_Uncertainity, 
      col.names = c ( "County", "Median Household Income(USD)", 
                      "Margin of Error(USD)", "MOE (%)", 
                      "Reliability Category"),
      caption = "Counties with Highest Income Estimate 
      Uncertainty in Pennsylvania (ACS 2022)"
      )
```

**Data Quality Commentary:**

[Write 2-3 sentences explaining what these results mean for algorithmic decision-making. Consider: Which counties might be poorly served by algorithms that rely on this income data? What factors might contribute to higher uncertainty?]

The counties like Forest, Sullivan, Union, Montour, and Elk with the highest margins of error are at risk of being misclassified by algorithms that rely heavily on median income estimates for resource allocation. **High uncertainty may be a result of smaller populations, sampling variability, or demographic complexity**, which can reduce survey precision. Policymakers should exercise caution when using these estimates for automated decision-making, as communities with unreliable data may be unfairly prioritized.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category

tibble (PA_Reliability) %>%
  arrange(desc(moe_pct))

selected_counties <- PA_Reliability %>%
  filter (NAME %in% c("Philadelphia County, Pennsylvania",
                     "Centre County, Pennsylvania",
                     "Forest County, Pennsylvania")) %>%
  select(NAME, county_name, totpopE, medincE, moe_pct, reliability) 

selected_counties
```

**Comment on the output:** The selected counties represent a range of data reliability levels and population sizes. Philadelphia County reflects challenges common in large, diverse urban areas. While Centre County represents a mid-sized county with moderate reliability, Forest County illustrates how small population size can influence margin of error and data confidence. Since there is no low confidence reliability, I have selected Forest county based on the it being closest to low confidence. Center County is selected based on relatively different characteristics to Philadelphia County.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names

# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter

# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations

# Add readable tract and county name columns using str_extract() or similar

acs_var1 <- c (white = "B03002_003", 
               black = "B03002_004", 
               hispanic = "B03002_012", 
               total_pop = "B03002_001")

counties <- c ("101", "027", "053") # Philadelphia, Centre, Forest

tract_data <- get_acs(geography = "tract", 
                      variables = acs_var1,
                      survey = "acs5",
                      year = 2022,
                      state = "PA",
                      county = counties,
                      output = "wide")

tract_data
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract

tract_data <- tract_data %>%
  mutate( white_pct = whiteE/total_popE*100,
          black_pct = blackE/total_popE*100,
          hisp_pct = hispanicE/total_popE*100,
          county_fips = str_sub(GEOID, 3, 5))

top_hisp_tract <- tract_data %>%
  arrange( desc(hisp_pct)) %>%
  slice(1) %>%
  select(NAME,hisp_pct)

top_hisp_tract

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group

county_demogrpahics <- tract_data %>%
  group_by(county_fips) %>%
  summarise (tracts = n(),
             avg_pct_white = mean(white_pct, na.rm = TRUE),
             avg_pct_black = mean(black_pct, na.rm = TRUE),
             avg_pct_hisp = mean(hisp_pct, na.rm = TRUE)
             )

# Create a nicely formatted table of your results using kable()
kable(county_demogrpahics, 
      col.names = c("County FIPS", "Tracts", "Avg. White %", "Avg. Black %", "Avg. Hispanic %"),
      caption = "Average Demographics by County for each Racial/Ethnic Group(ACS 2022)",
      digits = 1)

```
**Comment on the output:** Census Tract 195.02 in Philadelphia County, Pennsylvania has the highest percentage of Hispanic Population at **92.03%**.

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)

tract_data <- tract_data %>%
  mutate(white_moe_pct = whiteM/whiteE*100,
         black_moe_pct = blackM/blackE*100,
         hisp_moe_pct = hispanicM/hispanicE*100,
         tot_moe_pct = total_popM/total_popE*100)

# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

demo_MOE_pct <- tract_data %>%
  mutate(high_moe = ifelse(
  white_moe_pct > 15 |
  black_moe_pct > 15 |
  hisp_moe_pct > 15,
  "High MOE",
  "Acceptable MOE"
  ))

# Create summary statistics showing how many tracts have data quality issues

tract_data_sum <- demo_MOE_pct %>%
  count(high_moe) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

tract_data_sum

```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages

pattern_summary <- demo_MOE_pct %>%
  group_by(high_moe) %>%
  summarize(
    num_tracts = n(),
    avg_population = round(mean(total_popE, na.rm = TRUE), 0),
    avg_pct_white = round(mean(white_pct, na.rm = TRUE), 1),
    avg_pct_black = round(mean(black_pct, na.rm = TRUE), 1),
    avg_pct_hispanic = round(mean(hisp_pct, na.rm = TRUE), 1)
  )

# Use group_by() and summarize() to create this comparison
# Create a professional table showing the patterns

kable(pattern_summary,
      col.names = c("Data Quality Category", "Number of Tracts", "Average Population", "Avg % White", 
                    "Avg % Black", "Avg % Hispanic"),
      caption = "Demographic Patterns for each group")

```

**Pattern Analysis:** Tracts with high margins of error tend to have smaller populations and higher proportions of minority residents, particularly Hispanic and Black populations. This suggests that data reliability issues are not randomly distributed, but instead are more common in communities that may already face structural disadvantages. **These patterns raise concerns that systems relying on census estimates may systematically underrepresent or mischaracterize the needs of these communities.**


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**
**Overall Pattern Identification: What are the systematic patterns across all your analyses?**
The analysis reveals consistent and systematic patterns in both data reliability and demographic distribution across Pennsylvania. County-level analysis shows substantial variation in margins of error in key socioeconomic indicators such as median household income. The smaller and less populous counties exhibiting higher uncertainty. Tract-level analysis further demonstrates that data quality issues intensify at finer geographic scales, particularly for demographic variables. Margins of error are not evenly distributed. Instead, they cluster in specific geographies and population groups indicating that uncertainty is a structural feature of the data.

**Equity Assessment: Which communities face the greatest risk of algorithmic bias based on your findings?**
Communities facing the greatest risk of algorithmic bias are census tracts with smaller populations and higher concentrations of racial/ethinic (Black and Hispanic) residents. These tracts are disproportionately flagged for high margins of error across multiple demographic variables. If an algorithmic system were to rely on these estimates without accounting for reliability, it could systematically under-identify need or misallocate resources away from communities that already experience social and economic disadvantage. As a result, data uncertainty is an important equity issue. It potentially reinforces th existing disparities in access to public services and funding.

**Root Cause Analysis: What underlying factors drive both data quality issues and bias risk?**
The observed data quality issues stem from structural limitations in census data collection like lower response rates, higher residential mobility, language barriers and under representation in historically marginalized communities. Smaller population sizes at the tract level amplify statistical uncertainty. At the same time, socioeconomic and demographic factors further contribute to higher margins of error. These same conditions that generate unreliable data are often correlated with greater social service needs. This creates a compounding risk where communities most in need are also those least accurately represented in the data used to inform policy decisions.

**Strategic Recommendations: What should the Department implement to address these systematic issues?**
To mitigate these risks, the Department should adopt a data-informed but equity-conscious approach to algorithmic decision-making. First, margins of error should be explicitly incorporated into any prioritization model, with safeguards that prevent exclusion based solely on uncertain estimates. Second, areas with consistently high data uncertainty should be flagged for supplemental qualitative review or targeted outreach. Third, the Department should consider combining census data with administrative records and local knowledge to improve reliability in high-risk communities. Finally, transparent documentation of data limitations and regular bias audits should be institutionalized to ensure that algorithmic tools support equitable and accountable policy outcomes.

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

recommendations_table <- PA_Reliability %>%
  select(County = county_name,
         Median_Income = medincE,
         MOE_Percentage = moe_pct,
         Reliability = reliability) %>%
  mutate(Algorithm_Recommendations = case_when(
    Reliability == "High Confidence" ~ "Safe for algorithmic decisions",
    Reliability== "Moderate Confidence" ~ "Use with caution – monitor outcomes",
    Reliability== "Low Confidence" ~ "Requires manual review or additional data"
    )
    )

recommendations_table

# Format as a professional table with kable()
kable(recommendations_table,
      col.names = c ("County", "Median Income", "MOE %", "Data Reliability", "Algorithm Recommendations"),
      caption = "Algorithm Recommendations for County Data",
      digits = 1
      )

```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** 
All counties other then Cameroon, Carbon, Elk, Forest, Greene, Montour, Snyder, Sullivan, Union and Warren are classified as **High Confidence** as they are appropriate for immediate inclusion in algorithmic decision-making systems. Their key socioeconomic indicators exhibit low margins of error. The relatively high reliability of income estimates in these counties reduces the risk of misclassification or biased prioritization. Algorithms operating in these contexts are more likely to produce stable and defensible outcomes, making them suitable for initial deployment and pilot testing.

2. **Counties requiring additional oversight:** 
Counties like Cameroon, Carbon, Elk, Forest, Greene, Montour, Snyder, Sullivan, Union and Warren have **Moderate Confidence** data and should be included cautiously, with active monitoring and safeguards in place. While estimates in these counties are generally usable, elevated margins of error increase the risk of inconsistent outcomes over time. The Department should implement routine validation checks, track algorithmic outputs against observed service needs, and adjust thresholds as new data becomes available to ensure that communities are not inadvertently disadvantaged.

3. **Counties needing alternative approaches:** 
Counties classified as **Low Confidence** pose the highest risk for algorithmic bias due to substantial data uncertainty. In these areas, reliance on automated systems alone could result in inaccurate prioritization or exclusion of high-need populations. The Department should adopt alternative strategies such as manual case review, targeted community outreach, supplemental surveys, integration of administrative and local data sources before using algorithmic tools to guide funding or service decisions.

## Questions for Further Investigation

1. How do margins of error for key indicators change over time, and are certain counties consistently classified as low confidence across multiple ACS years? How are the margin of errors updated over time to improve reliability?

2. How does incorporating margins of error directly into algorithmic weighting schemes affect funding allocations and equity outcomes compared to traditional point-estimate models?

3. Are data reliability issues spatially clustered within specific regions of Pennsylvania, such as rural areas or post-industrial communities?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on February 3rd, 2026

**Reproducibility:** 
- All analysis conducted in R version [your version]
- Census API key required for replication
- Complete code and documentation available at: https://chavanindrajeet.github.io/IC_PPARep/

**Methodology Notes:**
**1.** The county selection for tract-level analysis was based on data reliability categories derived from margins of error, which introduces a subjective element in defining thresholds for “high,” “moderate,” and “low” confidence. While these thresholds follow common practice, alternative cutoffs could yield different county groupings. 

**2.** The demographic percentages and margins of error were calculated using ACS 5-year estimates, which represent pooled data across multiple years and may not reflect short-term changes. 

**3.** The tracts with zero or very small population estimates were implicitly affected by percentage and MOE calculations, potentially inflating uncertainty measures. 

**4.** The analysis relies on a single ACS year and excludes supplemental administrative data, meaning results are sensitive to data availability and census survey limitations.

**Limitations:**
Although no counties or census tracts fall into the low-reliability category, several areas exhibit moderate reliability with areas that are close to the threshold for low reliability. These areas may not fully represent the actual conditions within Philadelphia.
---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ Y ] All code chunks run without errors
- [ Y ] All "[Fill this in]" prompts have been completed
- [ Y ] Tables are properly formatted and readable
- [ Y ] Executive summary addresses all four required components
- [ Y ] Portfolio navigation includes this assignment
- [ Y ] Census API key is properly set 
- [ Y ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`