---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Indrajeet Chavan"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

In this lab, you will apply the spatial predictive modeling techniques demonstrated 
in the class exercise using a different 311 service request type of your choice as the predictor variable. You will build a complete spatial predictive model, document your process, and interpret your results.

### Learning Objectives

By completing this assignment, you will demonstrate your ability to:

- Adapt example code to analyze a new dataset
- Build spatial features for predictive modeling
- Apply count regression techniques to spatial data
- Implement spatial cross-validation
- Interpret and communicate model results
- Critically evaluate model performance

---

## Step 1: Choose Your 311 Violation Type

### Getting the Data and Requirements for Your Choice

After visiting the Chicago 311 Service Requests dataset, I have choosen a **different** violation type than the abandoned cars. 
The violations chosen are **Graffiti Removal and Vacant & Abandoned Buildings Reported**.

---

## Step 2: Complete the Analysis

Using the class exercises I have adapted the code to analyze your chosen 311 violation type.

### Overview of the assignment:

Work through all major sections of the analysis:

### Part 1: Data Loading & Exploration
Tasks in this section -
- Load your 311 data and Chicago spatial boundaries
- Create visualizations showing the spatial distribution of your violation type
- Describe patterns you observe

---

```{r}
#| message: false
#| warning: false

rm(list=ls())

# Load required packages
library(tidyverse)      # Data manipulation
library(tidycensus)
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals
library(dplyr) 

# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

# Create consistent theme for visualizations
theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

#### Exercise 1.1: Load Chicago Spatial Data

```{r}
# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)
policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")
```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
- It minimizes distortion in this region
- Uses feet (common in US planning)
- Allows accurate distance calculations

---

#### Exercise 1.2: Load Burglary Data and 311 Data

```{r}
#| message: false

# Load from provided data file (downloaded from Chicago open data portal)
burglaries <- st_read(here("Data", "burglaries.shp")) %>% 
  st_transform('ESRI:102271')

glimpse(burglaries)
glimpse(vacant)
glimpse(graffiti)

# Check the data
cat("\n✓ Loaded burglary data\n")
cat("  - Number of Burglaries:", nrow(burglaries), "\n")
cat("  - Number of Vacant & Abandoned Buildings:", nrow(vacant), "\n")
cat("  - Number of Graffiti Violations:", nrow(graffiti), "\n")
cat("  - CRS:", st_crs(burglaries)$input, "\n")
cat("  - Date range:", format(min(burglaries$Date, na.rm = TRUE)), "to", 
    format(max(burglaries$Date, na.rm = TRUE)), "\n")

```

**Takeaways:**
Question 1.1: How many burglaries are in the dataset? What time period does this cover? Why does the coordinate reference system matter for our spatial analysis?

Your answer here:
Number of Burglaries: 7482
Number of Vacant & Abondoned Buildings: 7482 
Number of Graffitti Violations: 7482 
Time period: 2017-01-01 to 2018-01-01
The coordinate reference system for Chicago matters because:
- It minimizes distortion in this region
- Uses feet (common in US planning)
- Allows accurate distance calculations

---

#### Exercise 1.3: Visualize Data

```{r}
#| fig-width: 10
#| fig-height: 5

# Simple point map
p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = burglaries, color = "#d62828", size = 0.1, alpha = 0.4) +
  labs(
    title = "Burglary Locations",
    subtitle = paste0("Chicago 2017, n = ", nrow(burglaries))
  )

# Density surface using modern syntax
p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(burglaries)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    guide = "none"  # Modern ggplot2 syntax (not guide = FALSE)
  ) +
  labs(
    title = "Density Surface",
    subtitle = "Kernel density estimation"
  )

# Combine plots using patchwork (modern approach)
p1 + p2 + 
  plot_annotation(
    title = "Spatial Distribution of Burglaries in Chicago",
    tag_levels = 'A'
  )

```

**Takeaways:**
Question 1.2: What spatial patterns do you observe? Are burglaries evenly distributed across Chicago? Where are the highest concentrations? What might explain these patterns?

Your answer here:

---

### Part 2: Fishnet Grid Creation
- Create a 500m x 500m fishnet grid
- Aggregate your violations to grid cells
- Visualize the count distribution

---

#### Exercise 

```{r}
# 2.1 Understanding the Fishnet
# Create 500m x 500m grid
# Create 500m x 500m grid
fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 500,  # 500 meters per cell
  square = TRUE
) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271') 

# Keep only cells that intersect Chicago
fishnet <- fishnet[chicagoBoundary, ]

# View basic info
cat("✓ Created fishnet grid\n")
cat("  - Number of cells:", nrow(fishnet), "\n")
cat("  - Cell size:", 500, "x", 500, "meters\n")
cat("  - Cell area:", round(st_area(fishnet[1,])), "square meters\n")

# 2.2 Aggregate violations to Grid
# Spatial join: which cell contains each burglary?
burglaries_fishnet <- st_join(burglaries, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countBurglaries = n()) 


glimpse(burglaries_fishnet)
st_crs(burglaries_fishnet)

# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(burglaries_fishnet, by = "uniqueID") %>%
  mutate(countBurglaries = replace_na(countBurglaries, 0))

# Summary statistics
cat("\nBurglary count distribution:\n")
summary(fishnet$countBurglaries)
cat("\nCells with zero burglaries:", 
    sum(fishnet$countBurglaries == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countBurglaries == 0) / nrow(fishnet), 1), "%)\n")

# 2.3 Visualize the data
# Visualize aggregated counts
ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Burglaries",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 1, 5, 10, 20, 40)
  ) +
  labs(
    title = "Burglary Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2017"
  ) +
  theme_crime()
```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Part 3: Create a Kernel Density Baseline
Before building complex models, let's create a simple baseline using Kernel Density Estimation (KDE).

The KDE baseline asks: "What if crime just happens where it happened before?" (simple spatial smoothing, no predictors)

#### Exercise 

```{r}
# Kernel Density for Burglaries
# Convert burglaries to ppp (point pattern) format for spatstat
burglaries_ppp <- as.ppp(
  st_coordinates(burglaries),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_burglaries <- density.ppp(
  burglaries_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster <- rast(kde_burglaries)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_value = terra::extract(
      kde_raster,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )

cat("✓ Calculated KDE baseline\n")

#Visulaize the data
ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_value), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of burglary locations"
  ) +
  theme_crime()
```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Part 4: Create Spatial Predictor Variables
Now we'll create features that might help predict burglaries. We'll use "broken windows theory" logic: signs of disorder predict crime.

#### Exercise: Load 311 Abandoned vacant plots and graffiti calls

```{r}
# Load 311 Vacant & Abondoned Buildings
vacant <- 
  read.csv("/Users/indrajeetchavan/Library/CloudStorage/OneDrive-PennO365/Penn/6. Penn Sem 2/CPLN 5920 PPA/GitHub/IC_PPARep/Labs/Labs_4/Data/311_Service_Requests_-_Vacant_and_Abandoned_Buildings_Reported_-_Historical_20260329.csv")

# Load 311 graffitti violations
graffiti <- 
  read.csv("/Users/indrajeetchavan/Library/CloudStorage/OneDrive-PennO365/Penn/6. Penn Sem 2/CPLN 5920 PPA/GitHub/IC_PPARep/Labs/Labs_4/Data/311_Service_Requests_-_Graffiti_Removal_-_Historical_20260408.csv")

graffiti <- graffiti %>%
  dplyr::select(Latitude, Longitude) %>%
  drop_na()

vacant <- vacant %>%
  dplyr::select(LATITUDE, LONGITUDE) %>%
  drop_na()

vacant <- vacant %>%
  rename(
    Latitude = LATITUDE,
    Longitude = LONGITUDE
  )

violations <- bind_rows(
  graffiti %>% mutate(type = "Graffiti"),
  vacant %>% mutate(type = "Vacant")
)

violations_sf <- st_as_sf(violations,
                         coords = c("Longitude", "Latitude"),
                         crs = 4326)

violations_sf <- st_transform(violations_sf, 'ESRI:102271')

vacant_sf <- violations_sf %>%
  filter(type == "Vacant")

graffiti_sf <- violations_sf %>%
  filter(type == "Graffiti")

cat("✓ Loaded abandoned vacant and abandoned buildings calls\n")
cat("  - Number of calls:", nrow(vacant), "\n")
cat("✓ Loaded graffiti violations calls\n")
cat("  - Number of calls:", nrow(graffiti), "\n")
```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Exercise: Count of Vacant & Abandoned Buildings and Graffiti per Cell

```{r}
# Aggregate violations to fishnet
violations_fishnet <- st_join(violations_sf, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(violations = n())

# Join to fishnet
fishnet <- fishnet %>%
  left_join(violations_fishnet, by = "uniqueID")

# fishnet$violations[is.na(fishnet$violations.y)] <- 0
# fishnet$violations[is.na(fishnet$violations.x)] <- 0

cat("Violation distribution:\n")
summary(fishnet$violations)

#creating the specific columns for visualization
violations_fishnet <- st_join(violations_sf, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID, type) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = type,
    values_from = count,
    values_fill = 0
  )

fishnet <- fishnet %>%
  left_join(violations_fishnet, by = "uniqueID")

names(fishnet)

ggplot() +
  geom_sf(data = fishnet, aes(fill = Graffiti), color = NA)

table(violations_sf$type)

#Visualization for all three types of violations
pa <- ggplot() +
  geom_sf(data = fishnet, aes(fill = Graffiti), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Graffiti Removal Requests") +
  theme_crime()

pb <- ggplot() +
  geom_sf(data = fishnet, aes(fill = Vacant), color = NA) +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Vacant Buildings") +
  theme_crime()

pc <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  scale_fill_viridis_c(option = "viridis") +
  labs(title = "Burglaries") +
  theme_crime()

(pa | pb | pc) +
  plot_annotation(title = "Comparing Urban Disorder and Crime Patterns")

```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Exercise: Nearest Neighbor Features

Count in a cell is one measure. Distance to the nearest 3 abandoned cars captures local context.

```{r}
# Get coordinates
fishnet_coords <- st_coordinates(st_centroid(fishnet))
violations_coords <- st_coordinates(violations_sf)

# Calculate k nearest neighbors (k = 3)
library(FNN)

#Check for NAs
sum(is.na(violations_coords))
sum(is.na(fishnet_coords))

#Clearing NAs
violations <- violations %>%
  drop_na(Latitude, Longitude)

violations_sf <- st_as_sf(violations,
                         coords = c("Longitude", "Latitude"),
                         crs = 4326)

#Remove NA Coordinates Before KNN
violations_coords <- st_coordinates(violations_sf)
violations_coords <- violations_coords[complete.cases(violations_coords), ]

## Clean violations coordinates ONLY
fishnet_coords <- st_coordinates(st_centroid(fishnet))
fishnet_coords <- fishnet_coords[complete.cases(fishnet_coords), ]

#Run Knn
nn_result <- get.knnx(violations_coords, fishnet_coords, k = 3)

# Add feature
fishnet$violations.nn <- rowMeans(nn_result$nn.dist)

# Add to fishnet
fishnet <- fishnet %>%
  mutate(
    violations.nn = rowMeans(nn_result$nn.dist)
  )

cat("✓ Calculated nearest neighbor distances\n")
summary(fishnet$violations.nn)

```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Exercise: Distance to Hot Spots

```{r}
# Function to calculate Local Moran's I
calculate_local_morans <- function(data, variable, k = 5) {
  
  # Create spatial weights
  coords <- st_coordinates(st_centroid(data))
  neighbors <- knn2nb(knearneigh(coords, k = k))
  weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
  
  # Calculate Local Moran's I
  local_moran <- localmoran(data[[variable]], weights)
  
  # Classify clusters
  mean_val <- mean(data[[variable]], na.rm = TRUE)
  
  data %>%
    mutate(
      local_i = local_moran[, 1],
      p_value = local_moran[, 5],
      is_significant = p_value < 0.05,
      
      moran_class = case_when(
        !is_significant ~ "Not Significant",
        local_i > 0 & .data[[variable]] > mean_val ~ "High-High",
        local_i > 0 & .data[[variable]] <= mean_val ~ "Low-Low",
        local_i < 0 & .data[[variable]] > mean_val ~ "High-Low",
        local_i < 0 & .data[[variable]] <= mean_val ~ "Low-High",
        TRUE ~ "Not Significant"
      )
    )
}

# Apply to burglaries
violations_burglary <- calculate_local_morans(
  violations_sf, 
  "burglaries",   # <-- change to your actual column name
  k = 5
)

# Apply to Vacant & Abandoned Buildings
violations_vacant <- calculate_local_morans(
  violations_sf, 
  "vacant_abandoned_buildings",  # adjust name if needed
  k = 5
)

# Apply to Graffiti Violations
violations_graffiti <- calculate_local_morans(
  violations_sf, 
  "graffiti_violations",  # adjust name if needed
  k = 5
)

# Visualize hot spots for Burglary
ggplot() +
  geom_sf(
    data = violations_burglary, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Burglary Clusters",
    subtitle = "High-High = Hot spots"
  ) +
  theme_crime()

# Visualize hot spots for Vacant & Abandoned Buildings
ggplot() +
  geom_sf(
    data = violations_vacant, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Burglary Clusters",
    subtitle = "High-High = Hot spots"
  ) +
  theme_crime()

# Visualize hot spots for Graffiti
ggplot() +
  geom_sf(
    data = violations_graffiti, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Burglary Clusters",
    subtitle = "High-High = Hot spots"
  ) +
  theme_crime()

# Hotspot Distances
#Burglaries
hotspots_burglary <- violations_burglary %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

if (nrow(hotspots_burglary) > 0) {
  violations_burglary <- violations_burglary %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(violations_burglary), hotspots_burglary %>% st_union())
      )
    )
}

#Vacant Buildings
hotspots_vacant <- violations_vacant %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

if (nrow(hotspots_vacant) > 0) {
  violations_vacant <- violations_vacant %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(violations_vacant), hotspots_vacant %>% st_union())
      )
    )
}

#Graffiti
hotspots_graffiti <- violations_graffiti %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

if (nrow(hotspots_graffiti) > 0) {
  violations_graffiti <- violations_graffiti %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(violations_graffiti), hotspots_graffiti %>% st_union())
      )
    )
}
```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Part 5: Spatial Cross-Validation (2017)
- Implement Leave-One-Group-Out cross-validation on 2017 data
- Calculate and report error metrics (MAE, RMSE)

#### Exercise: Distance to Hot Spots

```{r}

```

**Takeaways:**
We're using ESRI:102271 (Illinois State Plane East, NAD83, US Feet). This is appropriate for Chicago because:
It minimizes distortion in this region
Uses feet (common in US planning)
Allows accurate distance calculations

---

#### Part 6: Model Evaluation
- Compare to KDE baseline

#### CHALLENGE TASK: Temporal Validation (2018) *not mandatory, only if you want very strong spatial analytics muscles*
- Download 2018 crimes https://data.cityofchicago.org/Public-Safety/Crimes-2018/3i3m-jwuy/about_data
- Be sure to filter for BURGLARIES (FORCED ENTRY only to match the 2017 data)
- Aggregate 2018 violations to the same fishnet grid
- Use your 2017 model to predict 2018 counts
- Calculate temporal validation metrics
- Compare spatial vs. temporal validation performance

---

## Step 3: Write Your Analysis

### Critical Requirement: Explain Each Step

**For each major section, you must explain in your own words:**

- **What** you are doing in that step
- **Why** this step is important for the analysis
- **What** you found or learned from the results

Do not simply copy text from the example or from your AI friend. 
Think about the purpose of each technique and articulate it in your own words.

---

## Step 4: Format Your Document

### Formatting Requirements

Your rendered HTML document should be **professional and easy to read**:

#### ✓ Clean Code
- Remove unnecessary code, comments, or debugging lines
- Keep only essential code chunks
- Use `code-fold: show` in your YAML header

#### ✓ Clear Structure
- Use headers to organize sections
- Include a table of contents
- Add your name and date

#### ✓ Readable Text
- Ensure all text renders properly (no broken markdown)
- Check that headers appear as headers (not plain text)
- Use proper markdown formatting

#### ✓ Quality Visualizations
- All plots should have titles and labels
- Maps should be readable
- Use consistent color schemes

#### ✓ Professional Presentation
- Proofread for typos
- Remove any "Your answer here" placeholders
- Make sure all code runs without errors
- Make IT NEAT. WE GOTTA LOOK GOOD HERE.

---

## Submission Checklist

Before you submit, verify that your document includes:

### Required Components

- [ ] **Introduction**: Brief description of your chosen 311 violation type and why you selected it
- [ ] **Data Exploration**: Maps and summary statistics of your violation data
- [ ] **Fishnet Grid**: Successfully created and visualized grid with aggregated counts
- [ ] **Spatial Features**: At least 2-3 spatial features calculated (k-NN, distance measures, etc.)
- [ ] **Local Moran's I**: Spatial autocorrelation analysis with interpretation
- [ ] **Count Regression**: Both Poisson and Negative Binomial models with comparison
- [ ] **Spatial Cross-Validation**: LOGO CV implemented on 2017 data with reported metrics
- [ ] **Temporal Validation**: 2018 data aggregated and tested with 2017 model(OPTIONAL.)
- [ ] **Validation Comparison**: Comparison of spatial vs. temporal validation performance
- [ ] **Model Comparison**: Your model vs. KDE baseline for 2018 predictions
- [ ] **Error Analysis**: Maps and discussion of where model performs well/poorly
- [ ] **Written Explanations**: Your own words explaining each step


### Technical Requirements

- [ ] Document renders to HTML without errors
- [ ] Code is clean and well-organized
- [ ] All visualizations display properly
- [ ] Headers and formatting render correctly
- [ ] File is named successfully posted to your website

---


